Distributional Trees and Forests

### 2017-10-30 pkg/disttree/inst/draft/draft.Rnw
added text and edited plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1671 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6e63b6f5b31dba9a5850a5784e714a0a48b629f6

\newpage
\section{Methodology}
\label{Methodology}
\subsection{Distributional Fit}
Fitting a distributional model to a set of observations without considering any covariates is a well known procedure which can for example be done by applying the maximum likelihood method. This is also the chosen approach in the methodology described in this section.\\
The goal is to fit a distributional model $D(Y, \theta)$ to a response variable $Y \in \mathcal{Y}$ for a \textit{k}-dimensional vector of distribution parameters $\theta \in \Theta$ in case of a \textit{k}-parametric distribution. The distribution family has to be specified in advance such that a log-likelihood function $\ell(\theta; Y)$ is provided. In our application of precipitation forecasting the response variable, which is the power transformed total precipitation amount within 24 hours, is expected to be normally distribution with mean/location parameter $\mu$ and standard deviation/scale parameter $\sigma$ and left censored with censoring point 0. Therefore, the corresponding log-likelihood function $\ell_{tp}$ with parameter vector $\theta = (\mu, \sigma)$ is
$$
\ell_{tp}(\mu, \sigma; Y) = 
\begin{cases}
    \log\left(\frac{1}{\sigma} \cdot \phi(\frac{Y - \mu}{\sigma}) \right), & \text{if } Y > 0\\
    \vspace{0.1cm}
    \log\left(\Phi(\frac{-\mu}{\sigma})\right), & \text{if } Y = 0
\end{cases}
$$
where $\phi$ and $\Phi$ are the probability density function and the distribution function of the standard normal distribution $\mathcal{N}(0,1)$.\\
With this specification the task of fitting a distributional model turns into the task of determining the distribution parameter $\theta$. This is done by the following maximization.
$$
\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
$$
where $\{y_i\}_{i=1,\ldots,n}$ are the observations of the response variable $Y$.\\
In that way, a whole distribution is specified with all its features, including location, scale and shape. However, fitting one global model to the whole data set might be too much of a generalization and therefore not represent the data and its features in a reasonable way. Moreover, if covariates are available it is desirable to include them into the model as they might provide important information. In our application on precipitation data a wide range of covariates are available from numerical esemble predictions considering precipitation, temperature, sea level pressure and other meteorological quantities which might be of interest/have an influence on the predictions. In particular, these covariates can be used to separate the data into more homogeneous subgroups. Doing so after fitting a global model to the whole data set and then fitting a local model to each subgroup should improve the model remarkably. This procedure of splitting a data set into subgroups can be done by applying a tree algorithm.

@@

\begin{figure}[t!]
\begin{center}
<<rain_Axams_pred, fig=TRUE, echo=FALSE>>=
plot(x = x, y = y1, type = "l", col = "red", 
     main = paste0("July ", pday, dayending), ylab = "density", 
     xlab = "total precipitation (power transformed)",
     ylim = c(0,max(y1, y2, y3, pm1, pm2, pm3) + 0.01),
     xlim = c(-1.5,8))

@@

To get another impression of the quality of the predictions made by the tested methods a probability integral transform (PIT) histograms is plotted in Figure \ref{pit_oosample} for out-of-sample predictions of each of the models. A perfectly fitting model would lead to a histogram with all bars being of height 1. Overall, it can be stated that all four model provide well fitting models while the forest seems to be the one being closest to the ideal case.

@@

  raster::image(tirol.dem, col = rev(gray.colors(100)),
                main="Stations in Tyrol", ylab = "latitude", xlab = "longitude")
  plot(tirol.gadm, add = TRUE)
  points(sp, pch = 21, bg = level_col_crps_ssc, col = 1)
  #points(sp, pch = 21, bg = col_crps_ssc, col = 1)
  #points(sp, pch = 21, bg = transp_col_crps_ssc, col = 1)
  legend("topleft", fill=c(pal["forest"], pal["gamlss"], pal["gamboostLSS"], pal["EMOS"]), legend = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS"), cex = 1)

@
\caption{\label{map}Map of all observations with corresponding CRPS skills score for predictions (Forest vs. GAM (prespecified) vs. GAM (boosted))}
\end{center}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
%\begin{subfigure}{0.5\textwidth}
<<rain_cross_axams_rmse, fig=TRUE, echo=FALSE>>=
boxplot(rain_rmse[,c(2,3,4)], ylab = "RMSE",
        names = c("Forest", "GAM (prespecified)", "GAM (boosted)")) 
@
%\end{subfigure}
\caption{\label{boxplot_rmse_cross}RMSE values of the 10x10 cross validation at the station Axams}
\end{center}
\end{figure}
%\hspace{-1cm}
\begin{figure}[h!]
\begin{center}
%\begin{subfigure}{0.5\textwidth}
<<rain_cross_axams_ll, fig=TRUE, echo=FALSE>>=
boxplot(rain_ll[,c(2,3,4)], ylim = c(-2.8,-1.3), ylab = "Log-likelihood",
        names = c("Forest", "GAM (prespecified)", "GAM (boosted)")) 
#boxplot(rain_ll)
@
%\end{subfigure}
%\caption{\label{boxplots_rain}Results 10x10 cross validation at the station Axams\\
%left: RMSE, right: log-likelihood}
\caption{\label{boxplot_cross_ll}Log-likelihood values of the 10x10 cross validation at the station Axams (extracted interval [-2.8,-1.3])}
\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
<<rain_all_rmse, fig=TRUE, echo=FALSE>>=
  matplot(t(rmse[,2:4]), type = "l", lwd = 1, 
          col = gray(0.5, alpha = 0.5), lty = 1, axes = FALSE, 
          xlab = "", ylab = "RMSE", xlim = c(0.5, 3.5))
  boxplot(rmse[,2:4], add = TRUE,
        names = c("Forest", "GAM (prespecified)", "GAM (boosted)")) 
@
\caption{\label{boxplot_rmse_all}RMSE for predictions over all stations}
\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
<<rain_all_ll, fig=TRUE, echo=FALSE>>=
  matplot(t(ll[,2:4]), type = "l", lwd = 1, col = gray(0.5, alpha = 0.5), 
          lty = 1, axes = FALSE, 
          xlab = "", ylab = "Log-likelihood", xlim = c(0.5, 3.5), ylim = c(-4,-1))
  boxplot(ll[,2:4], add = TRUE,
        names = c("Forest", "GAM (prespecified)", "GAM (boosted)")) 
@
\caption{\label{boxplot_ll_all}Log-likelihood vallues for predictions over all stations (extracted interval [-4,-1])}
\end{center}
\end{figure}

>>>>

\newpage
\section{Methodology}
\label{Methodology}
\subsection{Distributional Fit}
Fitting a distributional model to a set of observations without considering any covariates is a well known procedure which can for example be done by applying the maximum likelihood method. This is also the chosen approach in the methodology described in this section.\\
The goal is to fit a distributional model $D(Y, \theta)$ to a response variable $Y \in \mathcal{Y}$ for a \textit{k}-dimensional vector of distribution parameters $\theta \in \Theta$ in case of a \textit{k}-parametric distribution. The distribution family has to be specified in advance such that a log-likelihood function $\ell(\theta; Y)$ is provided. In our application of precipitation forecasting the response variable, which is the power transformed total precipitation amount within 24 hours, is expected to be normally distribution with mean/location parameter $\mu$ and standard deviation/scale parameter $\sigma$ and left censored with censoring point 0. Therefore, the corresponding log-likelihood function $\ell_{tp}$ with parameter vector $\theta = (\mu, \sigma)$ is
$$
\ell_{tp}(\mu, \sigma; Y) = 
\begin{cases}
    \log\left(\frac{1}{\sigma} \cdot \phi(\frac{Y - \mu}{\sigma}) \right), & \text{if } Y > 0\\
    \vspace{0.1cm}
    \log\left(\Phi(\frac{-\mu}{\sigma})\right), & \text{if } Y = 0
\end{cases}
$$
where $\phi$ and $\Phi$ are the probability density function and the distribution function of the standard normal distribution $\mathcal{N}(0,1)$.\\
With this specification the task of fitting a distributional model turns into the task of determining the distribution parameter $\theta$. This is done by the following maximization.
$$
\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
$$
where $\{y_i\}_{i=1,\ldots,n}$ are the observations of the response variable $Y$.\\
In that way, a whole distribution is specified with all its features, including location, scale and shape. However, fitting one global model to the whole data set might be too much of a generalization and therefore not represent the data and its features in a reasonable way. Moreover, if covariates are available it is desirable to include them into the model as they might provide important information. In our application on precipitation data a wide range of covariates are available from numerical ensemble predictions considering precipitation, temperature, sea level pressure and other meteorological quantities which might be of interest/have an influence on the predictions. In particular, these covariates can be used to separate the data into more homogeneous subgroups. Doing so after fitting a global model to the whole data set and then fitting a local model to each subgroup should improve the model remarkably. This procedure of splitting a data set into subgroups can be done by applying a tree algorithm.

@@

\begin{figure}[t!]
\begin{center}
<<rain_Axams_pred, fig=TRUE, echo=FALSE>>=
plot(x = x, y = y1, type = "l", col = "red", 
     main = paste0("July ", pday, dayending), ylab = "Density", 
     xlab = "Total precipitation (power transformed)",
     ylim = c(0,max(y1, y2, y3, pm1, pm2, pm3) + 0.01),
     xlim = c(-1.5,8))

@@

To get another impression of the quality of the predictions made by the tested methods a probability integral transform (PIT) histograms is plotted in Figure \ref{pit_oosample} for out-of-sample predictions of each of the models. A perfectly fitting model would lead to a histogram with all bars being of height 1 as indicated by the red horizontal line. Overall, it can be stated that all four model provide well fitting models while the forest seems to be the one being closest to the ideal case.

@@

  raster::image(tirol.dem, col = rev(gray.colors(100)),
                main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude")
  plot(tirol.gadm, add = TRUE)
  points(sp, pch = 21, bg = level_col_crps_ssc, col = 1)
  #points(sp, pch = 21, bg = col_crps_ssc, col = 1)
  #points(sp, pch = 21, bg = transp_col_crps_ssc, col = 1)
  legend("topleft", fill=c(pal["forest"], pal["gamlss"], pal["gamboostLSS"], pal["EMOS"]), legend = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS"), cex = 1)

@
\caption{\label{map}Map of all observations with corresponding CRPS skills score for predictions}
\end{center}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
%\begin{subfigure}{0.5\textwidth}
<<rain_cross_axams_rmse, fig=TRUE, echo=FALSE, width=10, height=7>>=
boxplot(rain_rmse[,c(2,3,4,6)], ylab = "RMSE",
        names = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS")) 
@
%\end{subfigure}
\caption{\label{boxplot_rmse_cross}RMSE values of the 10x10 cross validation at the station Axams}
\end{center}
\end{figure}
%\hspace{-1cm}
\begin{figure}[h!]
\begin{center}
%\begin{subfigure}{0.5\textwidth}
<<rain_cross_axams_ll, fig=TRUE, echo=FALSE, width=10, height=7>>=
boxplot(rain_ll[,c(2,3,4,6)], ylim = c(-2.8,-1.3), ylab = "Log-likelihood",
        names = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS")) 
#boxplot(rain_ll)
@
%\end{subfigure}
%\caption{\label{boxplots_rain}Results 10x10 cross validation at the station Axams\\
%left: RMSE, right: log-likelihood}
\caption{\label{boxplot_cross_ll}Log-likelihood values of the 10x10 cross validation at the station Axams (extracted interval [-2.8,-1.3])}
\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
<<rain_all_rmse, fig=TRUE, echo=FALSE, width=10, height=7.4>>=
  matplot(t(rmse[,c(2,3,4,6)]), type = "l", lwd = 1, 
          col = gray(0.5, alpha = 0.5), lty = 1, axes = FALSE, 
          xlab = "", ylab = "RMSE", xlim = c(0.5, 4.5))
  boxplot(rmse[,c(2,3,4,6)], add = TRUE,
        names = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS")) 
@
\caption{\label{boxplot_rmse_all}RMSE for predictions over all stations}
\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
<<rain_all_ll, fig=TRUE, echo=FALSE, width=10, height=7.4>>=
  matplot(t(ll[,c(2,3,4,6)]), type = "l", lwd = 1, col = gray(0.5, alpha = 0.5), 
          lty = 1, axes = FALSE, 
          xlab = "", ylab = "Log-likelihood", xlim = c(0.5, 4.5), ylim = c(-4,-1))
  boxplot(ll[,c(2,3,4,6)], add = TRUE,
        names = c("Forest", "GAM (prespecified)", "GAM (boosted)", "EMOS")) 
@
\caption{\label{boxplot_ll_all}Log-likelihood values for predictions over all stations (extracted interval [-4,-1])}
\end{center}
\end{figure}
